### 2022-08-22 draft/draft.tex
still writing

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\setcitestyle{numbers,square,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}

@@

\begin{document}

@@

\setlength{\parindent}{10.0pt}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{remark}
\newtheorem{remark}{Remark}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}

@@

Graph and network data have become increasingly widespread in various
fields including sociology, neuroscience, biostatistics, and computer
science. This has resulted in challenges for researchers who rely on
traditional statistical and machine learning methods that are
incompatible with graph data and instead assume that the data exist as
feature vectors. To illustrate this, consider the typical approach to
building a statistical or machine learning model. Data are often
represented as an \(n \times p\) matrix
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\) in which
each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features
and each column is a set of \(n\) feature measurements. An analysis task
for these data might be to come up with a classification model
\(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical
values of each feature of a vector. For instance, \(\phi(x)\) might
first compute the distance from \(x\) to one of \(K\) points in
\(\mathbb{R}^p\) and assign \(x\) to the label of the nearest point.
Examples of this include linear discriminant analysis (in the case of
supervised learning) and Lloyd's algorithm or Gaussian mixture models
(in the case of unsupervised learning). However, this type of approach
is incompatible with data that are represented as graphs, in which each
observation is not a vector of numerical features but a set of
relationships to other observations. Instead of feature vector
\(x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top\),
we observe
\(a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}\) where
each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).

A more formal description of graph data is as follows: Suppose we
observe a network of \(n\) objects and pairwise relationships between
them. This network is represented by a graph object \(G = (V, E)\) with
vertex set \(V\), representing the objects, and edge set \(E\),
representing the pairwise relationships. The numeric representation of
these data is in the form of \emph{affinity matrix}
\(A \in \mathbb{R}^{n \times n}\) in which each \(A_{ij}\) is object
\(i\)'s relationship to object \(j\). We assume that the entries of
\(A\) represent affinities or similarities, i.e., the higher the value
of \(A_{ij}\), the stronger the relationship \(i\) has to \(j\). If
\(A_{ij} = 0\), then \(i\) has no direct relationship to \(j\). \(A\) is
symmetric if it represents an undirected graph in which the relationship
from \(i\) to \(j\) is the same as the relationship from \(j\) to \(i\).
\(A\) is binary, i.e., \(A \in \{0, 1\}^{n \times n}\), if it represents
an unweighted graph in which edges either exist or don't exist. If \(A\)
is binary, we call it an \emph{adjacency matrix}.

@@

Given a sample or dataset, a typical analysis task is statistical
inference, or the estimation of various parameters assuming the data
come from a random distribution or process. These estimated parameters
are often then used for making predictions or deriving insights about
the population. For example, when fitting a Gaussian mixture model, the
data are first assumed to come from a mixture of Gaussians. The model
fitting process then involves estimating the means and standard
deviations of each Gaussian component, along with the mixture weights.
The resulting model provides insight into where each mixture component
is located, how disperse each component is, and how the data are
distributed between the components, as well as a prediction indicating
to which mixture a new observation belongs. In order to perform a
similar type of analysis for graphs, we must first define probability
distributions from which such data can be sampled.

Within the scope of this dissertation, we focus primarily on unweighted
and undirected graphs without self-loops, with a brief discussion on
generalizing these methods to weighted or directed graphs. The adjacency
matrix that describes these graphs is binary, symmetric, and hollow. In
this setting, a plausible model is to sample each edge independently
from a Bernoulli distribution, i.e., \(A_{ij} \indep P_{ij}\) for some
\(P_{ij} \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\)
since \(A\) is symmetric, and \(A_{ii} = 0\) since \(A\) is hollow).
Then similar to how the edges are compiled into an adjacency matrix
\(A\), the edge probabilities can be compiled into an edge probability
matrix \(P \in [0, 1]^{n \times n}\). This type of graph model is
defined as a \emph{Bernoulli graph}.

\begin{definition}[Bernoulli graph]
\end{definition}

Statistical inference is not possible on a general Bernoulli graph with
arbitrary edge probabilities since the number of parameters is equal to
the number of observations. Additional structure must be introduced. One
such structured Bernoulli graph model is the \(\ER\) graph, which is
defined as follows.

\begin{definition}[$\ER$ graph]
\end{definition}

\hypertarget{block-models-for-community-detection}{%
\subsection{Block Models for Community
Detection}\label{block-models-for-community-detection}}

\hypertarget{the-stochastic-block-model}{%
\subsubsection{The Stochastic Block
Model}\label{the-stochastic-block-model}}

\hypertarget{generalizations-of-the-stochastic-block-model-the-degree-corrected-block-model-and-the-popularity-adjusted-block-model}{%
\subsubsection{Generalizations of the Stochastic Block Model: the Degree
Corrected Block Model and the Popularity Adjusted Block
Model}\label{generalizations-of-the-stochastic-block-model-the-degree-corrected-block-model-and-the-popularity-adjusted-block-model}}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{float}
\usepackage{mathtools}
\usepackage{natbib}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\setcitestyle{year,comma}
\usepackage{verbatim}
\usepackage{amsthm}
\usepackage{comment}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{apalike}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

@@

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{conjecture}{Conjecture}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{definition}
\newtheorem{hypothesis}{Hypothesis}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

@@

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\setlength{\parindent}{10.0pt}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
\newcommand{\SBM}{\mathrm{SBM}}
\newcommand{\DCBM}{\mathrm{DCBM}}
\newcommand{\argmin}{\mathrm{argmin}}

@@

Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science.
This has resulted in challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors.
To illustrate this, consider the typical approach to building a statistical or machine learning model.
Data are often represented as an \(n \times p\) matrix \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\) in which each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features and each column is a set of \(n\) feature measurements.
An analysis task for these data might be to come up with a classification model \(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical values of each feature of a vector \(x_i\) to calculate a predicted label \(z_i \in \{1, 2, ..., K\}\).
For instance, \(\phi(x)\) might first compute the distances from \(x\) to \(K\) points in \(\mathbb{R}^p\) and then assign \(x\) to the label of the nearest point.
Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm \citep{1056489} or Gaussian mixture models \citep{doi:10.1198/016214502760047131} (in the case of unsupervised learning).
However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object:
Instead of feature vector \(x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top \in \mathbb{R}^p\), we observe \(a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}^\top \in \mathbb{R}^n\) in which each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).
For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called \emph{graph embedding} or \emph{spectral clustering} \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.

\begin{example}
$K$-means clustering \citep{MacQueen1967} is a nonparametric clustering method that minimizes the objective function
\begin{equation}
\label{eq:kmeans}
W(x_1, ..., x_n) = \sum_{k = 1}^K \sum_{x_i : z_i = k} \|x_i - \bar{x}_k\|_2^2,
\end{equation}
in which $x_1, ..., x_n \in \mathbb{R}^p$ are feature vectors of a sample, $z_1, ..., z_n \in \{1, ..., K\}$ are cluster assignments we wish to optimize for, and $\bar{x}_1, ..., \bar{x}_K \in \mathbb{R}^p$ are the centroids of each cluster, which act as nuisance parameters. 
A popular implementation of $K$-means clustering is Lloyd's algorithm \citep{1056489}, which is a type of coordinate descent algorithm in which at each iteration labels are updated according to which centroid they are closest to and centroids are updated via the sample mean of each cluster.

If instead of vectors in $\mathbb{R}^p$ we have a graph, it is not obvious how to translate $K$-means clustering to these data. 
In particular, there is no inherent notion of centroid or sample mean for graph data. 
However, there are inherent notions of distances among vectors, such as shortest path distance, expected commute time, or resistance distance. 
One way to adapt $K$-means to graph data is to use an equivalent objective function as Eq.~\ref{eq:kmeans} that doesn't use centroids (which are just nuisance parameters) and is stated only in terms of distances:  
$$
\tilde{W}(x_1, ..., x_n) = \sum_{k=1}^K \sum_{x_i, x_j : z_i = z_j = k} d(x_i, x_j)^2.
$$
Here, $d(x_i, x_j)$ is the distance between objects $x_i$ and $x_j$. 
If the data live in a Euclidean vector space, then $d(x_i, x_j) = \|x_i - x_j\|_2$, 
or if the data are represented as a graph, then $d(x_i, x_j)$ represents some notion of graph distance. 
While this formulation of $K$-means clustering takes care of the lack of centroids, the implementation still requires some thought since Lloyd's algorithm includes the computation of centroids. 
An alternative algorithm is MacQueen's exchange algorithm \citep{MacQueen1967}, which involves cycling through each vertex and then, for each vertex, cycling through the labels and choosing the label that minimizes the objective function.

Yet another approach to adapting $K$-means clustering to graph data is to embed the graph to Euclidean space and then apply Lloyd's algorithm on the embedding. 
\end{example}

We now provide a more formal description of graph data:
Suppose we observe a network of \(n\) objects and pairwise relationships between them.
This network is represented by a graph \(G = (V, E)\) with vertex set \(V = \{v_1, ..., v_n\}\), representing the \(n\) objects, and edge set \(E\), representing the up to \(n (n - 1) / 2\) pairwise relationships (assuming that there are no self-loops).
The numeric representation of these data is in the form of \emph{affinity matrix} \(A \in \mathbb{R}^{n \times n}\) in which each \(A_{ij}\) represents object \(i\)'s relationship to object \(j\).
We assume that the entries of \(A\) represent affinities or similarities, i.e., the higher the value of \(A_{ij}\), the stronger the relationship \(i\) has to \(j\).
If \(A_{ij} = 0\), then \(i\) has no direct relationship to \(j\).
\(A\) is symmetric if it represents an undirected graph in which the relationship from \(i\) to \(j\) is the same as the relationship from \(j\) to \(i\).
\(A\) is binary, i.e., \(A \in \{0, 1\}^{n \times n}\), if it represents an unweighted graph in which edges either exist or don't exist.
If \(A\) is binary, we call it an \emph{adjacency matrix}.

\begin{example}
\label{ex:uk-faculty}
\citet{PhysRevE.77.016107} constructed a friendship network among $81$ faculty from various schools at a university in the UK. 
These data are represented as a graph in which each vertex is a faculty member and edges between pairs of vertices indicate whether the two faculty are friends. 
The corresponding adjacency matrix $A \in \{0, 1\}^{81 \times 81}$ has zeros along the diagonal since there are no self-loops, and $A_{ij} = A_{ji} = 1$ if it is observed that the $i^{th}$ and $j^{th}$ faculty members are friends. 
The following is a visualization of this graph, with the vertices labeled by school affiliation.

\begin{figure}[H]

{\centering \includegraphics{draft_files/figure-latex/friendship-network-1}

}

\caption{Friendship network of 81 faculty at a UK university. The vertices are labeled by school affiliation.}\label{fig:friendship-network}
\end{figure}
\end{example}

@@

Given a sample or dataset, a typical analysis task is statistical inference, or the estimation of various parameters under the assumption that the data come from a random distribution or process.
These estimated parameters are often then used for making predictions or deriving insights about the population.
For example, when fitting a Gaussian mixture model, the data are first assumed to come from a mixture of Gaussians.
The model fitting process then involves estimating the means and standard deviations of each Gaussian component, along with the mixture weights.
The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components, as well as a prediction indicating to which mixture a new observation belongs.
In order to perform a similar type of analysis for graphs, we must first define probability distributions from which such data can be sampled.

Within the scope of this work, we focus primarily on unweighted and undirected graphs without self-loops, with a brief discussion on generalizing these methods to weighted or directed graphs.
The adjacency matrix that describes these graphs is binary, symmetric, and hollow.
In this setting, a plausible model is to sample each edge independently from a Bernoulli distribution, i.e., \(A_{ij} \indep \Bernoulli(P_{ij})\) for some \(P_{ij} \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\) since \(A\) is symmetric, and \(A_{ii} = 0\) since \(A\) is hollow).
Then similar to how the edges are compiled into an adjacency matrix \(A\), the edge probabilities can be compiled into an edge probability matrix \(P \in [0, 1]^{n \times n}\).
This type of graph model is defined as a \emph{Bernoulli graph} (also called an inhomogeneous \(\ER\) graph).
If \(A\) is the adjacency matrix of a Bernoulli graph with edge probability matrix \(P\), we denote \(A \sim \BG(P)\).
If the vertices and edge probabilities are sampled as a sequence of \(n\) i.i.d. random variables from probability distribution \(F\), then we denote \(A \sim \BG(F, n)\).

Statistical inference is not possible on a general Bernoulli graph with arbitrary edge probabilities since the number of parameters (individual edge probabilities) is equal to the number of observations (presence or absence of an edge).
Additional structure must be introduced.
One such structured Bernoulli graph model is the \(\ER\) graph, first proposed by \cite{Gilbert:1959}, which is defined as follows:

\begin{definition}[$\ER$ graph]
\label{def:erdos-renyi}
Let $P$ be an $n \times n$ matrix such that each $P_{ij} \equiv \theta \in [0, 1]$ is a constant. 
Then $A \sim \BG(P)$ is an $\ER$ graph. 
\end{definition}

\begin{example}[Maximum likelihood estimator for the $\ER$ graph]
One possible estimator for the lone parameter of an $\ER$ graph, $\theta$, is the maximum likelihood estimator, which is found by maximizing the log-likelihood function $\ell(\theta; A) = \log \theta \sum\limits_{i < j} A_{ij} + \log (1 - \theta) \sum\limits_{i < j} (1 - A_{ij})$. 
This can be solved directly by setting $\frac{d \ell}{d \theta} = 0$, which yields $\hat{\theta} = \frac{2 |E|}{n (n-1)}$.
\end{example}

\(\ER\) graphs lie on the opposite end of the spectrum in that they restrict the model to a single parameter, which does not result in a very interesting model or reflect many networks observed in real data.
Much work has been done in developing various Bernoulli graph models that strike a balance between structure and flexibility.
Two common types of structure for graph models, that are of particular interest in this work, are edge probabilities based on latent communities \citep{doi:10.1080/0022250X.1971.9989788, NIPS2008_3578, Karrer_2011, 307cbeb9b1be48299388437423d94bf1}, called \emph{block models}, and edge probabilities based on positions in a latent space \citep{10.1007/978-3-540-77004-6_11, rubindelanchy2017statistical}, called \emph{latent space models}.

\textbf{{[}Maybe insert something here about the exponential family of random graph models and how, like mixture distributions, block models do not fit within this family?{]}}

\begin{example}
Consider a graph in which the vertices represent a collection of $n$ planets with intelligent life sending out radio waves, and the existence of an edge between vertices $i$ and $j$ represents whether planets $i$ and $j$ have established contact.  
Since radio waves decay according to the inverse-square law, a plausible model for the edge probability between two vertices is 
$$P_{ij} = \frac{C \omega_i \omega_j}{d_{ij}^2}$$ 
where $\omega_i$ is the $i^{th}$ planet's radio signal strength, $d_{ij}$ is the Euclidean distance between planets $i$ and $j$, and $C$ is a normalizing constant. 
Let $P$ be the $n \times n$ matrix of these edge probabilities. 
Then $A \sim \BG(P)$ represents the network of planets that have made contact.

To estimate $\omega = \begin{bmatrix} \omega_1 & \cdots & \omega_n \end{bmatrix}^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization. 
For simplicity, set $C = 1$. 
Then the log-likelihood function can be written as:
$$
\ell(\omega) = 
\sum_{i < j} A_{ij} \log \bigg( \frac{\omega_i \omega_j}{d_{ij}^2 -
\omega_i \omega_j} \bigg) + \log (d_{ij}^2 - \omega_i \omega_j) + const.
$$
The partial derivatives with respect to each $\omega_i$ are:
$$
\frac{\partial \ell}{\partial \omega_i} = \sum_{j \neq i} \frac{A_{ij}}{\omega_i} - \frac{\omega_j (1 - A_{ij})}{d_{ij}^2 - \omega_i \omega_j}.
$$
Setting the gradient to zero yields $n$ equations with $n$ unknowns, but it is not entirely obvious how to solve this system of equations. 
Alternatively, we can use gradient ascent or some other numerical optimization method, since this particular log-likelihood function is concave. 
\end{example}

\hypertarget{contributions-of-this-work}{%
\subsection{Contributions of this work}\label{contributions-of-this-work}}

In this work, we explore three types of block models, the stochastic block model, the degree corrected block model, and the popularity adjusted block model, as well as a family of latent space models called generalized random dot product graphs.
The contributions of this work are as follows.
First, we show that, similar to how the stochastic block model and degree corrected block model are generalized random dot product graphs with specific latent structures, the popularity adjusted block model is also a generalized random dot product graph with a specific structure.
Using estimators with well-established consistency properties for the generalized random dot product graph, we develop consistent community detection and parameter estimation algorithms for the popularity adjusted block model.

\hypertarget{sec:blockmodel}{%
\section{Block Models for Community Detection}\label{sec:blockmodel}}

Network analysis is often concerned with community detection.
This has motivated statisticians to develop random graph models with inherent community structure in which the probability of an edge between vertices \(i\) and \(j\) depend on the communities to which vertices \(v_i\) and \(v_j\) belong.
More formally, one assigns each vertex \(v_i\) a community label \(z_i\) and assumes a Bernoulli graph in which \(P_{ij} = g(z_i, z_j, \theta_i, \theta_j)\) for some function \(g: \{1, ..., K\}^2 \times \mathbb{R}^d \times \mathbb{R}^d \to [0, 1]\) and \(\theta_i \in \mathbb{R}^d\) is a vector of real-valued parameters associated with each vertex.
Within the context of this work, we will call such models \emph{block models}.
This type of model frmaework allows us to frame the goal of community detection as a statistical inference problem:
identify the true community labels, up to permutation of the labels.

In this section, we explore the three main types of block models:
the stochastic block model \citep{doi:10.1080/0022250X.1971.9989788}, the degree corrected block model \citep{Karrer_2011}, and the popularity adjusted block model \citep{307cbeb9b1be48299388437423d94bf1}.
The main focus is on how these models are connected, particularly in how they are nested models \citep{Noroozi2022}, as well as community detection and parameter estimation via likelihood maximization.
We will later return to these block models and the relationship to another class of random graph models in sections \ref{sec:grdpg} and \ref{sec:pabm-grdpg}.

\hypertarget{sec:sbm}{%
\subsection{The Stochastic Block Model}\label{sec:sbm}}

The stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788} is the simplest of the three block models and assumes that each pair of communities has a constant edge probability \(\theta_{k \ell}\).
We now give the formal definition of the SBM within the context of Bernoulli graphs:

\begin{definition}[Stochastic block model]
\label{def:sbm}
Let $G = (V, E)$ be a Bernoulli graph with $n$ vertices, described by random adjacency matrix $A$. 
Let $K \geq 1$ be an integer and $\theta_{k \ell} \in [0, 1]$ for each $k, \ell \in \{1, ..., K\}$ (if $G$ is undirected, then $\theta_{k \ell} = \theta_{\ell k}$). 
Let $z_1, ..., z_n \in \{1, ..., K\}$ be community labels associated with each vertex. 
If the edge probability matrix $P$ for this graph is such that 
$P_{ij} = \theta_{z_i, z_j}$ and $A \sim \BG(P)$, then $G$ is a stochastic block model. 
\end{definition}

There are a few conventions for exactly how a graph is sampled as an SBM.
Within the context of this work, we assume that for a particular SBM, the number of communities, \(K\), and the community edge probabilities, \(\theta_{k \ell}\), are fixed.
Then some possible ways of sampling from an SBM are as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Assume the size of the graph \(n\) is fixed.
  Then each vertex \(v_i\) has a label \(z_i\), and the edge probability matrix \(P\) is fixed.
\item
  We sample the vertices and edges in sequence and suppose that each vertex \(v_i\) comes with a label \(z_i\), which determines its edge probabilities to the other vertices.
\item
  The vertices are sampled in sequence, and each \(z_i\) is sampled as \(z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)\), \(\sum_k^K \alpha_k = 1\). The edge probabilities are then determined by the sample \(\{z_i\}\) and the parameters \(\{\theta_{k \ell}\}\)
\end{enumerate}

An extension to these sampling schema include Bayesian models, e.g., each \(\theta_{k \ell}\) is sampled from a distribution with support \([0, 1]\).

\begin{example}[Stochastic block model with $K = 2$ communities]
\label{ex:assort-sbm}
We construct an SBM with two communities in which $\theta_{11} = 1/2$, $\theta_{12} = 1/8$, and $\theta_{22} = 1/4$. 
Note that in this example, the within-community edge probabilities are greater than the between-community edge probability. 
A realization of this graph with $n_1 = n_2 = 32$ is illustrated by figure \ref{fig:assort-sbm}.

\begin{figure}[H]

{\centering \includegraphics{draft_files/figure-latex/assort-sbm-1}

}

\caption{Two-community stochastic block model with $\theta_{11}=1/2$, $\theta_{22}=1/4$, and $\theta_{12} = \theta_{21} = 1/8$.}\label{fig:assort-sbm}
\end{figure}
\end{example}

In most conceptions of graph models, the within-community edge probability is greater than the between-community edge probability, as in example \ref{ex:assort-sbm}.
For instance, in the UK faculty friendship network in example \ref{ex:uk-faculty}, it is natural to expect that faculty within the same school are more likely to be friends than faculty from two different schools, and a cursory look at the visualization of the graph appears to confirm this.
However, SBMs are not restricted to this type of community structure.
We could just as easily construct an example in which \(\theta_{11}\) and \(\theta_{22}\) are less than \(\theta_{12}\). Within the context of the SBM, we say that a graph is assortative if \(P\) is positive semidefinite and disassortative otherwise, which roughly correspond to communities with higher within-community edge probabilities and higher between-community edge probabilities, respectively.
For example, the SBM in example \ref{ex:assort-sbm} is assortative and has edge probability matrix \(P\) that is rank 2 with nonzero eigenvalues \(n (\frac{3}{8} \pm \frac{\sqrt{2}}{8}) > 0\).
In the following example, we describe a network which can be modeled as a disassortative SBM.

\begin{example}[Dating network as a disassortative stochastic block model]
\label{ex:sbm-dating}
Consider an undirected graph in which each vertex $v_i$ represents users of an online dating service, each label $z_i$ represents the user's gender, and each edge represents a successful match between pairs of users. 
For simplicity, we will restrict the labels to female and male (denoted as $1$ and $2$). 
If we model this as an SBM, then $\theta_{11}$ is the probability of a match between two female users, $\theta_{22}$ is the probability of a match between two male users, and $\theta_{12} = \theta_{21}$ is the probability of a match between a female user and a male user. Based on trends in the United States \citep{gallup-lgbt}, 
a plausible set of edge probabilities is $\theta_{11} = \theta_{22} = 0.02$ and $\theta_{12} = 0.2$, which are used in the sample visualized in figure \ref{fig:dating-sbm}, along with $\alpha_1 = \alpha_2 = 1/2$ and $n = 64$.

\begin{figure}[H]

{\centering \includegraphics{draft_files/figure-latex/dating-sbm-1}

}

\caption{Stochastic block model of a dating network. This model is disassortative.}\label{fig:dating-sbm}
\end{figure}
\end{example}

\begin{example}[Assortativity and disassortativity of the two-community SBM]
Let $P$ be the $n \times n$ probability matrix of a two-community SBM. 
Then $rank(P) = 2$, and its two nonzero eigenvalues are:

$$
\lambda = \frac{n}{2} \bigg( 
  \theta_{11} + \theta_{22} \pm 
  \sqrt{(\theta_{11} - \theta_{22})^2 + 4 \theta_{12}^2} 
\bigg).
$$
The first eigenvalue is always positive. 
The second eigenvalue is positive if $\theta_{11} \theta_{22} > \theta_{12}^2$. 
Thus, $P$ describes an assortative SBM if $\theta_{11} \theta_{22} > \theta_{12}^2$ or 
a disassortative SBM if $\theta_{11} \theta_{22} < \theta_{12}^2$. 
\end{example}
